Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-22 latex/main.tex
Add Muller cite
Closes #759

\section*{INTRODUCTION}

%%% No subheadings, please. Use of abbreviations should be kept to a minimum,
%%% and nonstandard abbreviations should be defined when first used in the text.

The genomic revolution has transformed how we study the evolution and spread of pathogens,
allowing real-time tracking of pathogens over time and space.
During the COVID-19 pandemic, thousands of SARS-CoV-2 genome sequences were generated daily,
quickly overwhelming typical phylogenetic approaches.
Although real-time phylogenetic methods now exist to cope with this flood of genomic data
(e.g., UShER \cite{Turakhia2021} and MAPLE \cite{DeMaio2023}),
we lack methods to build phylogenies that can integrate recombination
in the face of ongoing sampling of infected cases.

UShER is a maximum parsimony method developed during the pandemic to infer phylogenies
from large numbers of densely sampled SARS-CoV-2 genome sequences \cite{Turakhia2021}.
UShER has generated phylogenies updated daily \cite{McBroome2021} and
has proven to be a powerful resource that successfully enabled the world
to keep up with an evolving disease (e.g., through the Nextstrain web portal \cite{Hadfield2018}).
These phylogenies are used to identify and monitor
potentially epidemiologically important ``Pango'' lineages \cite{Rambaut2020,McBroome2024},
which might have acquired features to become Variants of Concern (VOC),
for example, increased transmissibility, disease severity, and immune escape.
Beyond supporting real-time genomic epidemiology,
UShER phylogenies have been used in a multitude of evolutionary and epidemiological studies of SARS-CoV-2
(e.g., \cite{Karthikeyan2022,Obermeyer2022,Turakhia2022,Bloom2023}).
UShER uses a tree to represent evolutionary change via mutation,
assuming no recombination \cite{Turakhia2021}.
However, given extensive evidence for recombination in SARS-CoV-2
\cite{Jackson2021,VanInsberghe2021,Focosi2022,Gutierrez2022,Ignatieva2022,Tamura2023},
the tree-like representation of UShER is no longer tenable.

@@

Sc2ts identified 12 Pango X lineages in the ARG as non-recombinants
(Table~\ref{tab:pango_x_lineages}), on the basis of the HMM and available data.
For all these Pango X lineages, the number of recombination-informative sites
fell below our threshold of detection (i.e. $k = 4$ in the HMM).
The fact that most of these Pango X lineages (except XAJ and XAN) differ from a single parent node in the ARG
by at most three mutations (Table~\ref{tab:pango_x_lineages}) strongly supports 
their origins as non-recombinants.
For example, the closely related XN and XAU can be explained by two mutations
and one mutation in the ARG, respectively (Figure~\ref{fig:subgraphs}C).
%%% TODO: Say something about XAJ and XAN.
%JK: we should weaken this statement a bit, as we're not certain that they 
% really are recombinants, especially XP.
However, sc2ts did not detect several Pango X lineages (XB, XP, and XAJ) as recombinants
because the HMM treated characteristic deletions as missing data.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{pango_x_subgraphs}
\caption{
Illustrative subgraphs in the ARG.
(A) Subgraph around the recombination node associated with XA.
Two "Group A" recombinant samples reported by Jackson et al. \cite{Jackson2021}) are shown.
(B) Subgraph around the recombination node associated with a group of ``nested'' Pango X lineages: we use the label Xx as a shorthand for this related group of recombinants; and
(C) Subgraph around the MRCA of the closely related non-recombinants XN and XAU.
The nodes assigned to a Pango X lineage are colored by that lineage.
The numbers inside the colored boxes indicate
the number of samples assigned to the Pango X lineage indicated by the same color.
The numbers below the vertical dashed lines indicate
the total number of samples within a collapsed clade.
For each recombination node, a genome schematic shows
the location of its inferred breakpoint interval relative to the Spike gene.
The ancestry above the parents of the recombination nodes or the MRCA is not shown for clarity.
For more detailed visualizations of these subgraphs, see Document S2.
}
\label{fig:subgraphs}
\end{figure}

@@

>>>>

\section*{INTRODUCTION}
% Classical phylo methods fundamentally do not apply to SARS-CoV-2 data, both
% in terms of the basic goal, assumptions about samples and how recombination
% can be treated.
Classical phylogenetics is largely concerned with inferring the evolutionary history 
of species~\cite{Felsenstein2004}, with 
% Not sure about this assertion! Yan?
samples separated by millions of years of evolution.
[This is rough outline text]
Complex models are required to fill in the gaps in our knowledge,
incorporating prior information about sequence evoltion etc etc.
In this context, recombination can be treated as a nuisance factor 
because we cannot hope to observe it over these timescales [or something
similar].
Phylogenetics has continued along this vein, 
although with various challenges on how to accomodate 
more closely related samples and to handle potential recombination met.

The data collected during the COVID-19 pandemic as part of global
surveillance fundamentally changed these assumptions. 
At the peak in [month] 202X, there were over X thousand samples 
per day deposited in GISAID\cite{Shu2017}, with a total of 
over 17 million as of Sep 2025. This superabundance of data 
overwhelmed classical phylogenetic methods, both in terms 
of the number of sequences that could be processed, but also
more fundamentally in terms of the underlying assumption of 
retrospective data sparsity. [TODO rephrase]
The UShER method \cite{Turakhia2021} conclusively demonstrated 
the value of simpler models that build a tree iteratively
in real-time, quickly becoming an indispensible element 
of the pandemic response\cite{Hinrichs2024}.

Recombination continues to be treated as a nuisance factor, however~\cite{Neches2020}. 
Overwhelming evidence demonstrates the existance of recombination in SARS-CoV-2
\cite{Jackson2021,VanInsberghe2021,Muller2022bayesian,Focosi2022,Gutierrez2022,Ignatieva2022,Tamura2023},
with multiple deep recombination events now known to be present in 
genealogy of the global population [cites??].
Despite this, phylogenies used as the basis for innumerable studies 
continue to depict the genealogy of SARS-CoV-2 as a single
tree~\cite{McBroome2021,Hadfield2018}.

% The genomic revolution has transformed how we study the evolution and spread of pathogens,
% allowing real-time tracking of pathogens over time and space.
% During the COVID-19 pandemic, thousands of SARS-CoV-2 genome sequences were generated daily,
% quickly overwhelming typical phylogenetic approaches.
% Although real-time phylogenetic methods now exist to cope with this flood of genomic data

% (e.g., UShER \cite{Turakhia2021} and MAPLE \cite{DeMaio2023}),
% we lack methods to build phylogenies that can integrate recombination
% in the face of ongoing sampling of infected cases.

% UShER is a maximum parsimony method developed during the pandemic to infer phylogenies
% from large numbers of densely sampled SARS-CoV-2 genome sequences \cite{Turakhia2021}.
% UShER has generated phylogenies updated daily \cite{McBroome2021} and
% has proven to be a powerful resource that successfully enabled the world
% to keep up with an evolving disease (e.g., through the Nextstrain web portal \cite{Hadfield2018}).
% These phylogenies are used to identify and monitor
% potentially epidemiologically important ``Pango'' lineages \cite{Rambaut2020,McBroome2024},
% which might have acquired features to become Variants of Concern (VOC),
% for example, increased transmissibility, disease severity, and immune escape.
% Beyond supporting real-time genomic epidemiology,
% UShER phylogenies have been used in a multitude of evolutionary and epidemiological studies of SARS-CoV-2
% (e.g., \cite{Karthikeyan2022,Obermeyer2022,Turakhia2022,Bloom2023}).
% UShER uses a tree to represent evolutionary change via mutation,
% assuming no recombination \cite{Turakhia2021}.

@@

Sc2ts identified 12 Pango X lineages in the ARG as non-recombinants
(Table~\ref{tab:pango_x_lineages}), on the basis of the HMM and available data.
For all these Pango X lineages, the number of recombination-informative sites
fell below our threshold of detection (i.e. $k = 4$ in the HMM).
The fact that most of these Pango X lineages (except XAJ and XAN) differ from a single parent node in the ARG
by at most three mutations (Table~\ref{tab:pango_x_lineages}) strongly supports 
their origins as non-recombinants.
For example, the closely related XN and XAU can be explained by two mutations
and one mutation in the ARG, respectively (Figure~\ref{fig:subgraphs}C), and we
believe the original recombinant designation to be based on erroneous data
(Supplementary document S1.XXX).
%%% TODO: Say something about XAJ and XAN.
%JK: we should weaken this statement a bit, as we're not certain that they 
% really are recombinants, especially XP.
However, sc2ts did not detect several Pango X lineages (XB, XP, and XAJ) as recombinants
because the HMM treated characteristic deletions as missing data.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\linewidth]{pango_x_subgraphs}
\caption{
Illustrative subgraphs in the ARG.
Boxes identify specific samples, colored by their Pango X classification;
smaller squares list extra samples descending from a node, hidden for brevity,
with all hidden samples taking the same Pango designation as their parent.
For the two recombination nodes, a genome bar above shows
the location of the inferred breakpoint interval relative to the Spike gene.
For simplicity, ancestral lineages above the recombination parents or the MRCA are omitted.
(A) Subgraph around the recombination node associated with XA.
Two "Group A" recombinant samples reported by Jackson et al. \cite{Jackson2021}) are shown.
(B) Subgraph around the recombination node associated with a group of ``nested'' Pango X lineages: we use the label Xx as a shorthand for this related group of recombinants; and
(C) Subgraph around the MRCA of the closely related non-recombinants XN and XAU.
For more detailed visualizations of these subgraphs, see Document S2.
}
\label{fig:subgraphs}
\end{figure}

@@

\subsection{Sc2ts recombination detection results for additional Pango X lineages}

\begin{table}
\centering
\caption{
Sc2ts recombination detection results for Pango X lineages not added to the ARG.
}
\label{tab:absent_pango_x}
\begin{tabular}{crrr}
\toprule
Pango & Samples in Viridian & Samples passing QC & Breakpoints \\
\midrule
XBJ & 2 & 1 & 1\\
XBP & 8 & 2 & 1\\
XBS & 19 & 15 & 1\\
XBW & 1 & 1 & 1\\
XCA & 11 & 4 & 1\\
XAK & 2 & 2 & 1\\
XAY & 15 & 11 & 6\\
XBC & 36 & 21 & 3\\
XBL & 2 & 2 & 0\\
XBT & 1 & 1 & 2\\
\bottomrule
\end{tabular}
\end{table}
